---
title: "hoopR"
author: "chad allison"
date: '2023-01-16'
output: github_document
---

### setup

```{r message = F, warning = F}
tictoc::tic() # start script runtime tracker
library(tidyverse)
library(hoopR)
theme_set(theme_minimal())
```

### data import

```{r warning = F}
# write_csv(load_mbb_pbp(seasons = 2023), "pbp_data.csv")
df = read_csv("pbp_data.csv", col_types = cols())
glimpse(df)
```

### getting end game data

```{r}
end_game = df |>
  filter(type_text == "End Game") |>
  select(home_team_name, away_team_name, home_score, away_score) |>
  mutate(win_team = ifelse(home_score > away_score, home_team_name, away_team_name),
         lose_team = ifelse(win_team == home_team_name, away_team_name, home_team_name)) |>
  rename(home_team = home_team_name, away_team = away_team_name)

sample_n(end_game, 10)
```

### getting team records

*note that only teams having played ten or more games are included*

```{r}
all_teams = sort(unique(c(end_game$home_team, end_game$away_team)))
team_records = data.frame(team = all_teams)

team_wins = end_game |>
  count(win_team) |>
  rename(team = win_team)

team_losses = end_game |>
  count(lose_team) |>
  rename(team = lose_team)

team_records = team_records |>
  left_join(team_wins, by = "team") |>
  rename(wins = n) |>
  left_join(team_losses, by = "team") |>
  rename(losses = n) |>
  mutate(wins = ifelse(is.na(wins), 0, wins),
         losses = ifelse(is.na(losses), 0, losses),
         games_played = wins + losses,
         win_prop = round(wins / games_played, 3)) |>
  filter(games_played >= 10)

all_teams = team_records$team
rm(team_wins, team_losses)

team_records |>
  arrange(desc(win_prop)) |>
  head(10)
```

### getting team offensive points per game stats

```{r}
team_off_ppg = data.frame(team = all_teams, off_ppg = NA)

get_off_ppg = function(team) {
  
  home_total = end_game |>
    filter(home_team == team) |>
    pull(home_score) |>
    sum()
  
  away_total = end_game |>
    filter(away_team == team) |>
    pull(away_score) |>
    sum()
  
  team_total = home_total + away_total
  team_gp = team_records$games_played[which(team_records$team == team)]
  team_ppg = round(team_total / team_gp, 3)
  return(team_ppg)
  
}

team_off_ppg = team_off_ppg |>
  mutate(off_ppg = sapply(team, get_off_ppg))

# top 10 teams, offensive ppg
team_off_ppg |>
  arrange(desc(off_ppg)) |>
  head(10)
```

### getting team defensive points per game stats

```{r}
team_def_ppg = data.frame(team = all_teams, def_ppg = NA)

get_def_ppg = function(team) {
  
  home_total = end_game |>
    filter(home_team == team) |>
    pull(away_score) |>
    sum()
  
  away_total = end_game |>
    filter(away_team == team) |>
    pull(home_score) |>
    sum()
  
  team_total = home_total + away_total
  team_gp = team_records$games_played[which(team_records$team == team)]
  team_ppg = round(team_total / team_gp, 3)
  return(team_ppg)
  
}

team_def_ppg = team_def_ppg |>
  mutate(def_ppg = sapply(team, get_def_ppg))

# top 10 teams, defensive ppg
team_def_ppg |>
  arrange(desc(def_ppg)) |>
  head(10)
```

### joining offensive and defensive points per game stats

```{r}
team_ppg = team_off_ppg |>
  left_join(team_def_ppg, by = "team")

team_ppg |>
  mutate(diff = off_ppg - def_ppg) |>
  arrange(desc(diff)) |>
  left_join(team_records, by = "team") |>
  ggplot(aes(diff, win_prop)) +
  geom_point(aes(col = win_prop)) +
  scale_color_gradient(high = "springgreen3", low = "indianred1") +
  labs(x = "offensive ppg - defensive ppg",
       y = "win percentage",
       title = "team win percentage by ppg differential") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")
```

### adding ppg stats to end game data and creating strength ratings

```{r}
end_expanded = end_game |>
  filter(home_team %in% all_teams & away_team %in% all_teams) |>
  left_join(team_ppg, by = c("home_team" = "team")) |>
  rename(home_off_ppg = off_ppg, home_def_ppg = def_ppg) |>
  left_join(team_ppg, by = c("away_team" = "team")) |>
  rename(away_off_ppg = off_ppg, away_def_ppg = def_ppg) |>
  mutate(ho_str = home_score - away_def_ppg,
         hd_str = away_off_ppg - away_score,
         ao_str = away_score - home_def_ppg,
         ad_str = home_off_ppg - home_score)

end_expanded |>
  pivot_longer(c(ho_str, hd_str, ao_str, ad_str), names_to = "stat", values_to = "value") |>
  mutate(stat = factor(stat, levels = c("ho_str", "hd_str", "ao_str", "ad_str"))) |>
  ggplot(aes(stat, value)) +
  geom_boxplot(aes(fill = stat), alpha = 0.75) +
  labs(fill = NULL, title = "boxplots of off/def strength ratings for all matchups") +
  theme(plot.title = element_text(hjust = 0.5))
```

**how to interpret this plot**: i have created these "strength" ratings which are intended to quantify how a team is performing relative to the opposing team's average performance. for example, `ho_str` is the difference between points scored by the home team and the opposing team's average points allowed per game. a positive value would indicate that the home team performed well offensively, and a negative value would indicate that the opposing team performed well defensively. of course, it isn't always as simple as positive value => offense good & negative value => defense good, but that is a simple way to interpret the stat.

i do not find it surprising that the values for `ho_str` and `hd_str` appear to be slightly higher than those of `ao_str` and `ad_str` because that would explain the home court advantage. as far as distributions go, the four statistics seem to be very similar aside from their difference in median. `hd_str` appears to have the highest median value, which corresponds to `ao_str` having the lowest median value






















### script runtime

```{r}
tictoc::toc()
```

































